Economic Theory of Fertility Explains Modern Natural Selection

### 2020-11-13 why-negative-selection.Rmd
Tweaks and literature

---
title: "Natural Selection in the 20th Century"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "`r Sys.Date()`"
abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. The strength of natural selection on a polygenic score can be predicted by its correlation with earnings and education. Effects are larger among low-income and less educated individuals, younger parents, and people not living with a partner. Effect sizes are larger when we correct for ascertainment bias. The direction of natural selection is reversed among older parents, or  controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
---

@@

Since UK Biobank subjects are not representative of the wider population, our
results suggest that naïve estimates of natural selection are at risk of
ascertainment bias. To correct for this, we weight participants using data from
the 2011 Census and from the 2006 General Household Survey (GHS). We try three
alternative weighting schemes: geographical weighting by census MSOA, age and
presence/absence of a partner; weighting by age and highest educational
qualification; and age, highest qualification, and age at first live birth, for
women only. Mean effect sizes across all polygenic scores are increased by 
`r mean_rel$Geographical` (geographical weighting), 
`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
further affected by weighting on other demographic variables. Since the Biobank
sample seems to be ascertained in ways that reduce estimates of selection
effects, we suspect that our weighted estimates remain conservative.

The UK Biobank data contains information on respondents' number of
siblings. Since respondents' polygenic scores are equal in expectation
to the mean scores of their parents, we can use this to look at selection 
effects in the parents' generation. We estimate equation (\ref{eq:regression})
using *number of siblings* (including the respondent) as the dependent variable.

The parents' generation has a further source of ascertainment bias: sampling
parents of respondents overweights parents who have many children. For instance,
parents of three children will have, on average, three times more children
represented in UK Biobank than parents of one child. Parents of no children will
by definition not be represented. To compensate, we reweight our preferred
weightings (Age/Qualification) by the inverse of *number of siblings*. We then
regress this on our polygenic scores. Selection effects are highly correlated
across the two generations, and most share the same sign (appendix Figure
\@ref(fig:plot-res-sibs)). Effect sizes appear larger in the parents'
generation. We treat this result cautiously, because when we split respondents
up by year of birth, we find few differences in effect sizes between early- and
late-born respondents, for either generation. In other words, since estimated
effect sizes change when we change the dependent variable, but do not change
over time within either dependent variable, the explanation may be due to
remaining ascertainment bias within our sample.

Although the direction of selection effects does not change between the generations,
there are some differences. In particular, after we residualize polygenic scores
on the first 100 principal components of the array data, selection effects on 
*number of siblings* are reduced by about ten percent on average, whereas effects on
*number of children* barely change (appendix Figure 
\@ref(fig:plot-regs-controlled-pcs)). One possible explanation for this is that
earlier fertility is driven more by poverty, which is geographically clustered
and therefore correlates with the broad-scale genetic variation captured by
principal components. (TODO: insurance motive.) Another difference is that
in the parents' generation, raw selection effects and effects controlling for
AFLB are positively correlated, whereas they are negatively correlated in
the respondents' generation.[^educ-exception] These differences suggest that
the mechanisms of natural selection may have changed over time, even though
the outcomes are broadly the same.

[^educ-exception] Educational attainment is an exception. It is selected 
against in the parents' generation, but selected for if we control for AFLB,
which is the same pattern as for respondents.

@@

This theory predicts that genetic variants will be selected for (or against) in
proportion to their effect on earnings. Figure
\@ref(fig:plot-cors-earnings-educ) plots selection effects on each PGS against
that PGS's correlation with earnings in a respondent's first job[^earnings-est],
and against its correlation with educational attainment, a predictor of lifetime
earnings. The raw relationships (left column) are strongly negative. If we plot
selection effects controlling for age at first live birth (right column), the
effect reverses and becomes positive. This is evidence that the labour market
plays an important role in natural selection, with substitution effects
dominating overall. We also re-estimate equation \@ref(eq:regression) for each
PGS, controlling for earnings in first job and education levels. Effect sizes
are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$, suggesting that
earnings and education indeed mediate the effect of PGS on fertility.

@@

@@

cors_income <- cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
# TODO: use age_left_edu when this includes college
# TODO:
cors_educ <-   cor(famhist[score_names], famhist$age_fulltime_edu, use = "pairwise")

@@

This theory can explain why selection effects have opposite signs when AFLB
is controlled for. Suppose that people are aware of their endowments of human
capital, and that education and innate human capital are complements. They
choose how long to stay in education. After leaving education they enter the
labour market and form families. If so, then people with less human capital will
leave education earlier and have their first child earlier. This is a pure
substitution effect: the income effect plays no role in the decision to leave
education, since people are not yet earning wages. Whilst in the labour
market, higher earners will have more children (i.e. income effects dominate).
These assumptions will lead to the pattern we observe: earnings-increasing
variants lead to later first births (and fewer children overall), but more
children for any given AFLB.

The theory can also explain why natural selection is weaker among higher-income
parents. @becker1976child show that if parents care about child quality as well
as quantity, under certain assumptions the income effect will be stronger at
higher income levels; this can lead to a U-shaped relationship between income
and fertility.

* TODO: more.

Lastly, it is often assumed that the substitution effect will dominate
for lone parents, or those in unstable relationships, since they have less
opportunity to share childcare responsibilities, while the income effect will
dominate for couples who are able to reap the gains from specialization. If so,
then earnings-increasing genetic variants may decrease fertility more among single 
parents than among couples, as in our data. The same idea could explain our
results for lifetime number of sexual partners, if this is associated with
relationship stability.

Previous work shows similar relationships between *earnings* or *education* and
fertility, to the relationships shown here between *genetics* and fertility.
Higher income is associated with lower fertility [@becker1960economic;
@jones2006economic].
@cohen2013financial find that income decreases fertility at low income levels
but increases it at higher income levels. In a US sample, fertility decreases
with education faster among single mothers than married mothers 
[@baudin2015fertility].

>>>>

---
title: "Economic Theory of Fertility Explains Modern Natural Selection"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "`r Sys.Date()`"
abstract: "Natural selection has been documented in contemporary humans, but little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. Polygenic scores which predict lower earnings and education are selected for. Effects are larger among low-income and less educated people, younger parents, and people not living with a partner. Overall effect sizes are substantially bigger when we correct for ascertainment bias. The direction of natural selection is reversed among older parents, or controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
---

@@

Since UK Biobank subjects are not representative of the wider population, our
results suggest that naïve estimates of natural selection are at risk of
ascertainment bias. To correct for this, we weight participants using data from
the 2011 Census and from the 2006 General Household Survey (GHS). We try three
alternative weighting schemes: geographical weighting by census MSOA, age and
presence/absence of a partner; weighting by age and highest educational
qualification; and age, highest qualification, and age at first live birth, for
women only. Mean effect sizes across all polygenic scores are increased by a
factor of `r mean_rel$Geographical` (geographical weighting), 
`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
further affected by weighting on other demographic variables. Since the Biobank
sample seems to be ascertained in ways that shrink estimates of selection
effects, we suspect that our weighted estimates are still conservative.

The UK Biobank data contains information on respondents' number of
siblings. Since respondents' polygenic scores are equal in expectation
to the mean scores of their parents, we can use this to look at selection 
effects in the parents' generation. We estimate equation (\ref{eq:regression})
using *number of siblings* (including the respondent) as the dependent variable.
The parents' generation has a further source of ascertainment bias: sampling
parents of respondents overweights parents who have many children. For instance,
parents of three children will have, on average, three times more children
represented in UK Biobank than parents of one child. Parents of no children will
by definition not be represented. To compensate, we reweight our preferred
weightings (Age/Qualification) by the inverse of *number of siblings*.

Selection effects are highly correlated across the two generations, and most
share the same sign (appendix Figure \@ref(fig:plot-res-sibs)). Effect size
estimates are larger for the parents' generation. We treat this result cautiously,
because when we split respondents up by year of birth, we find few differences
in effect sizes between early- and late-born respondents, for either generation.
In other words, since estimated effect sizes change when we change the dependent
variable, but do not change over time within either dependent variable, this
may be due to remaining ascertainment bias within the sample.

Although the direction of selection effects does not change between the generations,
there are other differences. After we residualize polygenic scores on the first
100 principal components of the array data, selection effects on *number of
siblings* are reduced by about ten percent on average, whereas effects on
*number of children* barely change (appendix Figure
\@ref(fig:plot-regs-controlled-pcs)). This could be because
earlier fertility is driven more by geographically clustered deprivation [e.g.
via an insurance motive, @rendall1998old], which may correlate with the
broad-scale genetic variation captured by principal components. Another
difference is that in the parents' generation, raw selection effects and effects
controlling for AFLB are positively correlated, whereas they are negatively
correlated in the respondents' generation.[^educ-exception] These differences
suggest that the mechanisms of natural selection may have changed over time,
even though the outcomes are broadly the same.

[^educ-exception]: Educational attainment is an exception. It is selected 
against in the parents' generation, but selected for if we control for AFLB,
which is the same pattern as for respondents.

@@

This theory predicts that genetic variants will be selected for (or against) in
proportion to their effect on earnings. Figure
\@ref(fig:plot-cors-earnings-educ) plots selection effects on each PGS against
that PGS's correlation with earnings in a respondent's first job[^earnings-est],
and against its correlation with educational attainment, a predictor of lifetime
earnings. The raw relationships (left column) are strongly negative. If we plot
selection effects controlling for age at first live birth (right column), the
effect reverses and becomes positive. Thus, the labour market appears to
play an important role in natural selection. Substitution effects
dominate income effects overall, in accordance with the widespread association 
between income and lower fertility [@becker1960economic;@jones2006economic].

@@

We also re-estimate equation \@ref(eq:regression) for each
PGS, controlling for earnings in first job and education levels. Effect sizes
are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$, suggesting that
earnings and education indeed mediate the effect of PGS on fertility.

@@

cors_income <- cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
# TODO: use age_left_edu when this includes college

cors_educ <-   cor(famhist[score_names], famhist$age_fulltime_edu, use = "pairwise")

@@

This theory can explain why selection effects have opposite signs when AFLB is
controlled for. Suppose that people are aware of their endowments of human
capital, and that education and innate human capital are complements. They
choose how long to stay in education. After leaving education they enter the
labour market and form families. If so, then people with less human capital will
leave education earlier and have their first child earlier [cf.
@caucutt2002women; @monstad2008education]. This is a pure substitution effect: the income effect plays
no role in the decision to leave education, since people are not yet earning
wages.[^no-borrow] Whilst in the labour market, higher earners will have more
children (i.e. income effects dominate). These assumptions will lead to the
pattern we observe: earnings-increasing variants lead to later first births (and
fewer children overall), but more children for any given AFLB.

[^no-borrow]: This argument assumes that capital market imperfections hinder
people from borrowing against future income.

The theory can also explain why natural selection is weaker among higher-income
parents. @becker1976child show that if parents care about child quality as well
as quantity, under certain assumptions the income effect will be stronger at
higher income levels; this can lead to a U-shaped relationship between income
and fertility. Empirically, @cohen2013financial find that income decreases 
fertility at low income levels but increases it at higher income levels.

Lastly, it is often assumed that the substitution effect will dominate for lone
parents, or those in unstable relationships, since they have less opportunity to
share childcare responsibilities, while the income effect will dominate for
couples who are able to reap the gains from specialization. Indeed, US fertility
decreases faster with education among single mothers than married mothers
[@baudin2015fertility].If so, then earnings-increasing genetic variants may
decrease fertility more among single parents than among couples, as in our data.
The same idea could explain our results for lifetime number of sexual partners,
if this is associated with relationship stability.
